---
title : "HW 4"
author : "Zhen Wu"
output: pdf_document
---

### Prob. 1

#### (a) Function to check if a value is above a threshold
```{r}
check_above <- function(x, y) {
  return(x > y)
}
```

#### (b) For loop using a random normal vector
```{r}
set.seed(42)
random_vec <- rnorm(n = 10, mean = 35, sd = 10)
loop_results <- logical(10)

for (i in 1:10) {
  loop_results[i] <- check_above(random_vec[i], 35)
}
print(random_vec)
print(loop_results)
```

#### (c) Using sapply on the airquality dataset
```{r}
# We use na.rm = TRUE because the airquality dataset contains NA values
ozone_check <- sapply(airquality$Ozone, check_above, y = 42)
ozone_above_42 <- sum(ozone_check, na.rm = TRUE)
cat("Number of days with Ozone > 42:", ozone_above_42, "\n")
```

#### (d) Using mapply for multiple variables simultaneously
```{r}
target_data <- list(airquality$Ozone, airquality$Temp, airquality$Solar.R)
thresholds <- c(42, 77.88, 185.9)

multi_counts <- mapply(function(col_data, thresh) {
  sum(sapply(col_data, check_above, y = thresh), na.rm = TRUE)
}, target_data, thresholds)

names(multi_counts) <- c("Ozone > 42", "Temp > 77.88", "Solar.R > 185.9")
print(multi_counts)
```

### Prob. 2
#### (a) Fibonacci Implementation
# Recursive approach
```{r}
fib_recursive <- function(n) {
  if (n <= 2) {
    return(1)
  } else {
    return(fib_recursive(n - 1) + fib_recursive(n - 2))
  }
}

# Iterative (loop) approach
fib_loop <- function(n) {
  if (n <= 2) return(1)
  fib_values <- numeric(n)
  fib_values[1:2] <- 1
  for (i in 3:n) {
    fib_values[i] <- fib_values[i - 1] + fib_values[i - 2]
  }
  return(fib_values[n])
}
```

#### (b) Benchmarking for n = 30
```{r}
# Measuring execution time for both methods
time_recursive <- system.time(fib_recursive(30))
time_loop <- system.time(fib_loop(30))

cat("Time for Recursion (n=30):\n")
print(time_recursive)

cat("\nTime for Loop (n=30):\n")
print(time_loop)
```

### Prob. 3

The construction of the $LU$ decomposition relies on the specific structure of the atomic lower triangular matrices $L_k = I - l_k e_k^T$. The first "stroke of luck" is found in the simplicity of the inverse of these matrices. Because the vector $l_k$ has zeros in its first $k$ entries, the inner product $e_k^T l_k$ is zero. This leads directly to the inverse:

$$L_k^{-1} = I + l_k e_k^T$$

The second stroke of luck involves the product of these inverses to form the final matrix $L$. When we multiply $L_1^{-1} L_2^{-1} \dots L_{m-1}^{-1}$, the "cross terms" of the form $l_i e_i^T l_j e_j^T$ vanish whenever $i < j$ because the $i$-th entry of $l_j$ is zero. This allows the product to simplify to a basic sum of the rank-one updates:

$$L = \prod_{k=1}^{m-1} (I + l_k e_k^T) = I + \sum_{k=1}^{m-1} l_k e_k^T$$

The third stroke of luck is the culmination of the algorithm itself. By applying these Gauss transformations sequentially to the matrix $A$, we obtain an upper triangular matrix $U$. The combined effect of these algebraic properties ensures that we can write the original matrix as a product of a unit lower triangular matrix and an upper triangular matrix:

$$A = LU$$

---

### Prob. 4

Let $R$ be a nonsingular $m \times m$ upper-triangular matrix and define the subspace $S_n = \text{span}\{e_1, \dots, e_n\}$. Since $R$ is upper-triangular, each column $r_j$ satisfies $r_j \in S_j$. For any $n$, the first $n$ columns span a subspace of $S_n$, and due to the nonsingularity of $R$, these columns are linearly independent. By a dimension argument, we have:

$$\text{span}\{r_1, \dots, r_n\} = S_n$$

To show that $X = R^{-1}$ is upper-triangular, we consider the equation $R x_j = e_j$ for the $j$-th column of the inverse. Since $e_j \in S_j$, and we know that $S_j$ is exactly the span of the first $j$ columns of $R$, the vector $e_j$ can be written as:

$$e_j = \sum_{i=1}^j c_i r_i = R \sum_{i=1}^j c_i e_i$$

Because $R$ is nonsingular, the solution $x_j$ is unique. The expression above implies that $x_j$ must be a linear combination of only the first $j$ standard basis vectors:

$$x_j = \begin{bmatrix} c_1 & \dots & c_j & 0 & \dots & 0 \end{bmatrix}^T$$

Since every column $x_j$ has zeros below the $j$-th entry, the inverse matrix $R^{-1}$ is upper-triangular.

---

### Prob. 5

For the induced 2-norm, we evaluate the maximum magnification of a unit vector $x$. Using the definition of the matrix norm and the property that $v^* x$ is a scalar, we have:

$$\|E\|_2 = \sup_{\|x\|_2=1} \|u v^* x\|_2 = \|u\|_2 \sup_{\|x\|_2=1} |v^* x|$$

By applying the Cauchy-Schwarz inequality, the maximum value of $|v^* x|$ for a unit vector $x$ is exactly $\|v\|_2$. Thus, the 2-norm of the outer product is the product of the Euclidean norms of the vectors:

$$\|E\|_2 = \|u\|_2 \|v\|_2$$

Regarding the Frobenius norm, we use the trace identity $\|E\|_F^2 = \text{Tr}(E^* E)$. Substituting $E = uv^*$ and using the cyclic property of the trace, we obtain:

$$\|E\|_F^2 = \text{Tr}(v u^* u v^*) = \text{Tr}(u^* u v^* v)$$

Since $u^* u = \|u\|_2^2$ and $v^* v = \|v\|_2^2$ are scalars, the trace of their product is simply the product itself:

$$\|E\|_F^2 = \|u\|_2^2 \|v\|_2^2$$

Taking the square root proves that $\|E\|_F = \|u\|_2 \|v\|_2$. Therefore, the relationship holds true for both the induced 2-norm and the Frobenius norm in the case of a rank-one matrix.