---
title: HW5
author: Zhen Wu
output: 
  pdf_document:
    pandoc_args: ["--katex"]
---

## Prob. 6
```{r}
mgs <- function(x) {
  n <- ncol(x)
  m <- nrow(x)
  q <- matrix(0, nrow = m, ncol = n)
  r <- matrix(0, nrow = n, ncol = n)

  for (j in seq_len(n)) {
    v <- x[, j]

    for (i in seq_len(j - 1)) {
      r[i, j] <- t(q[, i]) %*% v
      v <- v - r[i, j] * q[, i]
    }

    r[j, j] <- sqrt(sum(v^2))
    q[, j] <- v / as.numeric(r[j, j])
  }

  q
}

cgs <- function(x) {
  n <- ncol(x)
  m <- nrow(x)
  q <- matrix(0, nrow = m, ncol = n)
  r <- matrix(0, nrow = n, ncol = n)

  for (j in seq_len(n)) {
    v <- x[, j]

    for (i in seq_len(j - 1)) {
      r[i, j] <- t(q[, i]) %*% x[, j]
      v <- v - r[i, j] * q[, i]
    }

    r[j, j] <- sqrt(sum(v^2))
    q[, j] <- v / as.numeric(r[j, j])
  }

  q
}

epsilon <- 1e-6
a1 <- c(1, epsilon, 0, 0)
a2 <- c(1, 0, epsilon, 0)
a3 <- c(1, 0, 0, epsilon)
M <- cbind(a1, a2, a3)

qcgs <- cgs(M)
qmgs <- mgs(M)

print(qcgs)
print(qmgs)
```

It can easily verified that the output of MGS has better orthogonality than that of CGS.

## Prob. 7
To prove the equivalence between the Modified Gram-Schmidt (MGS) method and the Householder QR factorization of the augmented matrix, we proceed by verifying the construction of the Householder reflections $P_k$ and the final transformation matrix $P$.

First, let $v_k = \begin{bmatrix} -e_k \\ q_k \end{bmatrix} \in \mathbb{R}^{m+n}$, where $q_1, \dots, q_n$ are the orthonormal vectors generated by the MGS process applied to $A$. We observe that 
$$
v_k^T v_k = (-e_k)^T(-e_k) + q_k^T q_k = 1 + 1 = 2
$$

A standard Householder reflection is defined as 
$$
H = I - 2 \frac{vv^T}{v^T v}
$$ 
Substituting our expression for $v_k$, we obtain 
$$
P_k = I - 2 \frac{v_k v_k^T}{2} = I - v_k v_k^T
$$
For any $k \neq j$, the inner product $v_k^T v_j = e_k^T e_j + q_k^T q_j = 0 + 0 = 0$, because the standard basis vectors $e_k$ and the MGS vectors $q_k$ are respectively orthogonal. 

Since these vectors are mutually orthogonal, the product of the Householder matrices simplifies to a summation: $P = P_1 P_2 \dots P_n = I - \sum_{k=1}^n v_k v_k^T$.Next, we evaluate the block structure of $P$. The summation term is given by:$$\sum_{k=1}^n v_k v_k^T = \sum_{k=1}^n \begin{bmatrix} -e_k \\ q_k \end{bmatrix} \begin{bmatrix} -e_k^T & q_k^T \end{bmatrix} = \sum_{k=1}^n \begin{bmatrix} e_k e_k^T & -e_k q_k^T \\ -q_k e_k^T & q_k q_k^T \end{bmatrix} = \begin{bmatrix} I_n & -Q^T \\ -Q & QQ^T \end{bmatrix}$$where $Q = [q_1, \dots, q_n]$. Subtracting this from the identity matrix $I_{m+n} = \begin{bmatrix} I_n & 0 \\ 0 & I_m \end{bmatrix}$, we find:$$P = \begin{bmatrix} I_n - I_n & Q^T \\ Q & I_m - QQ^T \end{bmatrix} = \begin{bmatrix} 0_n & Q^T \\ Q & I_m - QQ^T \end{bmatrix}$$This confirms the block form of $P$ stated in the hint.Finally, we show that applying $P^T$ (which equals $P$ since it is symmetric) to the augmented matrix yields the upper triangular matrix $R$. Let $A = QR$ be the QR factorization of $A$ via MGS. Then:$$P^T \begin{bmatrix} 0_n \\ A \end{bmatrix} = \begin{bmatrix} 0_n & Q^T \\ Q & I - QQ^T \end{bmatrix} \begin{bmatrix} 0_n \\ QR \end{bmatrix} = \begin{bmatrix} Q^T (QR) \\ (I - QQ^T) QR \end{bmatrix} = \begin{bmatrix} (Q^T Q) R \\ QR - Q(Q^T Q) R \end{bmatrix}$$Since the columns of $Q$ are orthonormal, $Q^T Q = I_n$. Substituting this into the equation results in:$$P^T \begin{bmatrix} 0_n \\ A \end{bmatrix} = \begin{bmatrix} I_n R \\ QR - QR \end{bmatrix} = \begin{bmatrix} R \\ 0 \end{bmatrix}$$This proves that the Householder QR factorization of the augmented matrix $\begin{bmatrix} 0_n \\ A \end{bmatrix}$ is indeed equivalent to the MGS method applied to $A$, as it produces the same upper triangular matrix $R$ and utilizes the MGS vectors $q_k$ within the reflections.

## Prob. 8
To verify the FLOP count for the Givens QR factorization of a matrix $A \in \mathbb{R}^{m \times n}$, we consider the cost of zeroing out each element below the main diagonal using successive Givens rotations. For each column $j$ from $1$ to $n$, we eliminate the elements $a_{ij}$ for $i = j+1, \dots, m$. A single Givens rotation $G(j, i, \theta)$ targeted at the element $(i, j)$ requires calculating the rotation parameters $c = \cos \theta$ and $s = \sin \theta$, which involves a constant number of operations ($O(1)$) that is negligible in the overall asymptotic count. The primary computational cost lies in applying the rotation to the two affected rows, $j$ and $i$, across the remaining columns $k = j+1, \dots, n$.

For each such column $k$, the transformation is defined by:
$$
\begin{bmatrix} a_{jk} \\ a_{ik} \end{bmatrix} \leftarrow \begin{bmatrix} c & -s \\ s & c \end{bmatrix} \begin{bmatrix} a_{jk} \\ a_{ik} \end{bmatrix}
$$
This update requires 4 multiplications and 2 additions per column, totaling $6$ FLOPs per index $k$. Since there are $n-j$ columns to update after the pivot column, the cost for one rotation is approximately $6(n-j)$ FLOPs. Summing this over all required rotations to clear column $j$, we obtain the total cost:
$$
\text{FLOPs} \approx \sum_{j=1}^n \sum_{i=j+1}^m 6(n-j) = \sum_{j=1}^n 6(m-j)(n-j)
$$
To simplify the algebra, we perform a change of variables by letting $k = n-j$. As $j$ ranges from $1$ to $n$, the new index $k$ ranges from $n-1$ down to $0$. We can rewrite the term $m-j$ as $(m-n) + (n-j) = (m-n) + k$. 

Substituting these into the summation, we get:
$$
6 \sum_{k=0}^{n-1} ((m-n) + k)k = 6 \left[ (m-n) \sum_{k=0}^{n-1} k + \sum_{k=0}^{n-1} k^2 \right]
$$
Using the standard power sum formulas $\sum_{k=1}^{N} k = \frac{N(N+1)}{2}$ and $\sum_{k=1}^{N} k^2 = \frac{N(N+1)(2N+1)}{6}$, we focus on the leading-order terms $N^2/2$ and $N^3/3$ where $N = n-1 \approx n$. The expression becomes:$$6 \left[ (m-n) \frac{n^2}{2} + \frac{n^3}{3} \right] = 3(m-n)n^2 + 2n^3$$Expanding the first term gives $3mn^2 - 3n^3$. Combining this with the $2n^3$ term results in the final leading-order count:
$$
\text{FLOPs} \approx 3mn^2 - n^3
$$

## Prob. 9

```{r}
# CGS
cgs_qr <- function(A) {
  m <- nrow(A)
  n <- ncol(A)
  Q <- matrix(0, m, n)
  R <- matrix(0, n, n)
  for (j in 1:n) {
    v <- A[, j]
    if (j > 1) {
      for (i in 1:(j - 1)) {
        R[i, j] <- sum(Q[, i] * A[, j])
        v <- v - R[i, j] * Q[, i]
      }
    }
    R[j, j] <- sqrt(sum(v^2))
    Q[, j] <- v / R[j, j]
  }
  return(list(Q = Q, R = R))
}

# MGS
mgs_qr <- function(A) {
  m <- nrow(A)
  n <- ncol(A)
  V <- A
  Q <- matrix(0, m, n)
  R <- matrix(0, n, n)
  for (j in 1:n) {
    R[j, j] <- sqrt(sum(V[, j]^2))
    Q[, j] <- V[, j] / R[j, j]
    if (j < n) {
      for (k in (j + 1):n) {
        R[j, k] <- sum(Q[, j] * V[, k])
        V[, k] <- V[, k] - R[j, k] * Q[, j]
      }
    }
  }
  return(list(Q = Q, R = R))
}

# Householder
householder_qr <- function(A) {
  m <- nrow(A)
  n <- ncol(A)
  R <- A
  Q <- diag(m)
  for (k in 1:n) {
    x <- R[k:m, k]
    e1 <- c(1, rep(0, length(x) - 1))
    v <- sign(x[1]) * sqrt(sum(x^2)) * e1 + x
    v <- v / sqrt(sum(v^2))


    R[k:m, k:n] <- R[k:m, k:n] - 2 * v %*% (t(v) %*% R[k:m, k:n])

    Q[, k:m] <- Q[, k:m] - 2 * (Q[, k:m] %*% v) %*% t(v)
  }
  return(list(Q = Q[, 1:n], R = R[1:n, 1:n]))
}

# Generate Vandermonde matrix
p <- seq(0, 1, length.out = 25)
A <- outer(p, 0:14, "^")

# Condition numbers
cond_A <- kappa(A, exact = TRUE)
cond_AtA <- kappa(t(A) %*% A, exact = TRUE)

# 3. Running decompositions and calculating errors
res_cgs <- cgs_qr(A)
res_mgs <- mgs_qr(A)
res_hh  <- householder_qr(A)

calc_errors <- function(A, res) {
  Q <- res$Q
  R <- res$R
  err_back <- sqrt(sum((A - Q %*% R)^2)) # ||A - QR||_F
  err_orth <- sqrt(sum((diag(ncol(Q)) - t(Q) %*% Q)^2)) # ||I - QtQ||_F
  return(c(BackErr = err_back, OrthErr = err_orth))
}

results <- rbind(
  CGS = calc_errors(A, res_cgs),
  MGS = calc_errors(A, res_mgs),
  Householder = calc_errors(A, res_hh)
)

print(list(Cond_A = cond_A, Cond_AtA = cond_AtA))
print(results)
```