---
title: HW6
author: Zhen Wu
output: pdf_document
---
## Prob. 10

Since $P$ is a nonzero projector, it satisfies the idempotency property by definition.

$$
P^2 = P
$$

Any eigenvalue $\lambda$ of a projector must satisfy the equation $\lambda^2 = \lambda$, which implies that the only possible eigenvalues are $\lambda = 0$ or $\lambda = 1$.

Because $P$ is given as a nonzero matrix, it must possess at least one eigenvalue equal to $1$.

Recall that the spectral radius $\rho(P)$, defined as the maximum absolute value of the eigenvalues of $P$, provides a lower bound for the induced $2$-norm of any matrix.

$$
\|P\|_2 \geq \rho(P)
$$

Substituting the known spectral radius into this inequality, we obtain the first part of the proof.

$$
\|P\|_2 \geq 1
$$

Now, we consider the condition for equality, $\|P\|_2 = 1$.

Suppose $P$ is an orthogonal projector, which implies $P$ is Hermitian (or symmetric in $\mathbb{R}^{m \times m}$), meaning $P = P^T$.

The singular values of $P$ are defined as the square roots of the eigenvalues of $P^T P$.

For an orthogonal projector, we have the following identity.

$$
P^T P = P P = P^2 = P
$$

Since the eigenvalues of $P$ are $1$ and $0$, the singular values of $P$ are $\sqrt{1} = 1$ and $\sqrt{0} = 0$.

The $2$-norm is the largest singular value, so $\|P\|_2 = 1$.

Conversely, suppose $\|P\|_2 = 1$.

Let the singular value decomposition (SVD) of $P$ be given by the following expression.

$$
P = U \Sigma V^T
$$

In this decomposition, $U$ and $V$ are orthogonal matrices, and $\Sigma$ is a diagonal matrix of singular values $\sigma_1 \geq \sigma_2 \geq \dots \geq \sigma_r > 0$.

Given $\|P\|_2 = 1$, we have $\sigma_1 = 1$, and consequently, $1 \geq \sigma_i > 0$ for all $i \in \{1, \dots, r\}$.

Substituting the SVD into the projector equation $P^2 = P$, we get the following.

$$
(U \Sigma V^T)(U \Sigma V^T) = U \Sigma V^T
$$

By multiplying by $U^T$ on the left and $V$ on the right, the equation simplifies.

$$
\Sigma (V^T U) \Sigma = \Sigma
$$

Let $W = V^T U$, where $W$ is an orthogonal matrix.

The equation can be written in terms of its components for the first $r$ indices.

$$
\sigma_i w_{ii} \sigma_i = \sigma_i \implies w_{ii} = \frac{1}{\sigma_i}
$$

Since $\sigma_i \leq 1$, it follows that $w_{ii} \geq 1$.

However, the entries of an orthogonal matrix must satisfy $|w_{ij}| \leq 1$, which forces $w_{ii} = 1$ and $\sigma_i = 1$ for all $i \in \{1, \dots, r\}$.

Because the diagonal entries $w_{ii}$ are all $1$, the orthogonality of $W$ requires that all other entries in the first $r$ rows and columns must be zero.

$$
W = \begin{pmatrix} I_r & 0 \\ 0 & W_{22} \end{pmatrix}
$$

This structure implies $U = V W$.

Substituting this back into the SVD of $P$ yields the following result.

$$
P = V W \Sigma V^T = V \begin{pmatrix} I_r & 0 \\ 0 & W_{22} \end{pmatrix} \begin{pmatrix} I_r & 0 \\ 0 & 0 \end{pmatrix} V^T = V \begin{pmatrix} I_r & 0 \\ 0 & 0 \end{pmatrix} V^T
$$

The resulting expression for $P$ is clearly symmetric.

$$
P^T = (V \Sigma V^T)^T = V \Sigma V^T = P
$$

Since $P$ is a symmetric projector, it is by definition an orthogonal projector.

## Prob. 11

To solve the problem of computing the smallest singular value for the given $m \times m$ matrix $A$, we first define the structure of the matrix and the mathematical relationship between its singular values and eigenvalues.

The matrix $A$ is defined as follows.

$$
A = \begin{pmatrix} 0.1 & 1 & \dots & 1 \\ 0 & 0.1 & \dots & 1 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & 0.1 \end{pmatrix} \in \mathbb{R}^{m \times m}
$$

The singular values $\sigma_i$ of $A$ are the square roots of the eigenvalues of the matrix $A^T A$.

$$
\sigma_{min}(A) = \sqrt{\lambda_{min}(A^T A)}
$$

The following R code implements both the direct SVD approach and the eigenvalue approach for $1 \le m \le 30$.

```{r}
m_values <- 1:30
svd_results <- numeric(30)
eigen_results <- numeric(30)

for (m in m_values) {
  A <- matrix(1, nrow = m, ncol = m)
  A[lower.tri(A)] <- 0
  diag(A) <- 0.1

  # Method 1: Direct SVD
  s_values <- svd(A)$d
  svd_results[m] <- min(s_values)
  # Method 2: Eigenvalues of A^T * A
  ata_eigen <- eigen(t(A) %*% A, symmetric = TRUE)$values
  # Ensure no negative values due to precision issues before sqrt
  eigen_results[m] <- sqrt(max(min(ata_eigen), 0))
}

# Plotting the results on a log scale
plot(m_values, svd_results, type = "b", col = "blue", log = "y", pch = 16,
     xlab = "Matrix size m", ylab = "Smallest Singular Value (log scale)",
     main = "Comparison of Smallest Singular Value Computation")
lines(m_values, eigen_results, type = "b", col = "red", pch = 4, lty = 2)
legend("bottomleft", legend = c("svd(A)", "sqrt(min(eigen(A^T A)))"),
       col = c("blue", "red"), pch = c(16, 4), lty = c(1, 2))
```

## Prob. 12

The following R code constructs ten random matrices according to the specified distribution and computes the spectral abscissa alongside the matrix exponential norm over the interval $t \in [0, 20]$.

The matrix $A$ is generated as $R - 2I$, where $R$ consists of standard normal entries.

The library expm is utilized for the computation of the matrix exponential.
```{r}
library(expm)

n <- 10
t_vals <- seq(0, 20, length.out = 200)
num_trials <- 10

plot(NULL, xlim = c(0, 20), ylim = c(1, 1e4), log = "y",
     xlab = "t", ylab = "Norm / Spectral Abscissa Bound",
     main = "Norm of exp(tA) vs exp(t * alpha(A))")

for (i in 1:num_trials) {
  A <- matrix(rnorm(n^2), nrow = n) - 2 * diag(n)

  # Calculate spectral abscissa alpha(A)
  eigs <- eigen(A, only.values = TRUE)$values
  alpha_A <- max(Re(eigs))

  # Compute ||exp(tA)|| and exp(t*alpha(A))
  norm_vals <- sapply(t_vals, function(t) norm(expm(t * A), type = "2"))
  bound_vals <- exp(t_vals * alpha_A)

  # Plot results
  lines(t_vals, norm_vals, col = "blue", lwd = 1)
  lines(t_vals, bound_vals, col = "red", lty = 2)
}

legend("topright", legend = c("||exp(tA)||", "exp(t * alpha(A))"),
       col = c("blue", "red"), lty = c(1, 2))
```

The spectral abscissa $\alpha(A)$ governs the long-term asymptotic behavior of the matrix exponential norm, though it often overlooks short-term dynamics.

$$
\alpha(A) = \max_{j} \text{Re}(\lambda_j)
$$

In non-normal matrices, where $A^T A \neq AA^T$, non-orthogonal eigenvectors can cause transient growth, leading the norm to spike significantly before eventually decaying.

$$
\| e^{tA} \|_2 \gg e^{t \alpha(A)} \text{ for small } t
$$

If the spectral abscissa is determined by a complex conjugate pair of eigenvalues, the curve will exhibit persistent oscillations as $t \to \infty$.

$$
\lambda = \alpha(A) \pm i\beta, \quad \beta \neq 0
$$

The imaginary component $\beta$ introduces sinusoidal fluctuations that prevent the norm from following a simple exponential path.

$$
\| e^{tA} \|_2 \propto e^{t \alpha(A)} \left| \cos(\beta t + \phi) \right|
$$

Conversely, if the dominant eigenvalue is purely real, the norm will eventually approach a smooth, monotonic exponential growth or decay.

$$
\lim_{t \to \infty} \frac{\| e^{tA} \|_2}{e^{t \alpha(A)}} = C
$$

## Prob. 13

To find the eigenvalue decomposition of the block matrix $B$, we define the eigenvalue problem for a partitioned vector.

$$
B \begin{bmatrix} x \\ y \end{bmatrix} = \lambda \begin{bmatrix} x \\ y \end{bmatrix}
$$

Expanding the matrix-vector product for the matrix $B = \begin{bmatrix} 0 & A^T \\ A & 0 \end{bmatrix}$, we obtain a system of coupled equations.

$$
\begin{bmatrix} 0 & A^T \\ A & 0 \end{bmatrix} \begin{bmatrix} x \\ y \end{bmatrix} = \begin{bmatrix} A^T y \\ Ax \end{bmatrix} = \begin{bmatrix} \lambda x \\ \lambda y \end{bmatrix}
$$

This system implies two primary relations: $A^T y = \lambda x$ and $Ax = \lambda y$.

By substituting the second equation into the first, we can decouple the variables and identify the relationship to the singular values of $A$.

$$
A^T (Ax) = A^T (\lambda y) = \lambda (A^T y) = \lambda^2 x
$$

Similarly, substituting the first into the second yields the following.

$$
A (A^T y) = A (\lambda x) = \lambda (Ax) = \lambda^2 y
$$

These results demonstrate that $x$ and $y$ must be eigenvectors of $A^T A$ and $AA^T$, respectively, with the shared eigenvalue $\lambda^2$.

From the provided singular value decomposition $A = U \Sigma V^T$, we know that $A^T A = V \Sigma^2 V^T$ and $AA^T = U \Sigma^2 U^T$.

The eigenvalues of $A^T A$ are the squares of the singular values $\sigma_i^2$, meaning the eigenvalues of $B$ satisfy $\lambda^2 = \sigma_i^2$, or $\lambda = \pm \sigma_i$.

The corresponding eigenvectors $x$ and $y$ are the right singular vectors $v_i$ and left singular vectors $u_i$.


For each singular value $\sigma_i > 0$, we can construct two eigenvectors for $B$.

If $\lambda = \sigma_i$ and we set $x = v_i$, then $y = \frac{1}{\sigma_i} A v_i = u_i$.

If $\lambda = -\sigma_i$ and we set $x = v_i$, then $y = -\frac{1}{\sigma_i} A v_i = -u_i$.

By normalizing these vectors, we obtain the orthonormal eigenvectors of $B$.

$$
q_i^{+} = \frac{1}{\sqrt{2}} \begin{bmatrix} v_i \\ u_i \end{bmatrix}, \quad q_i^{-} = \frac{1}{\sqrt{2}} \begin{bmatrix} v_i \\ -u_i \end{bmatrix}
$$

We can now assemble the full eigenvalue decomposition $B = Q \Lambda Q^T$.

The $2m \times 2m$ diagonal matrix of eigenvalues $\Lambda$ contains the singular values and their negatives.

$$
\Lambda = \begin{bmatrix} \Sigma & 0 \\ 0 & -\Sigma \end{bmatrix}
$$

The corresponding orthogonal matrix $Q$ is formed by concatenating the normalized eigenvectors.

$$
Q = \frac{1}{\sqrt{2}} \begin{bmatrix} V & V \\ U & -U \end{bmatrix}
$$

The orthogonality of $Q$ is verified by the following product.

$$
Q^T Q = \frac{1}{2} \begin{bmatrix} V^T & U^T \\ V^T & -U^T \end{bmatrix} \begin{bmatrix} V & V \\ U & -U \end{bmatrix} = \frac{1}{2} \begin{bmatrix} V^T V + U^T U & V^T V - U^T U \\ V^T V - U^T U & V^T V + U^T U \end{bmatrix} = \begin{bmatrix} I & 0 \\ 0 & I \end{bmatrix}
$$

Therefore, the eigenvalue decomposition of matrix $B$ is given by the following expression.

$$
B = \left( \frac{1}{\sqrt{2}} \begin{bmatrix} V & V \\ U & -U \end{bmatrix} \right) \begin{bmatrix} \Sigma & 0 \\ 0 & -\Sigma \end{bmatrix} \left( \frac{1}{\sqrt{2}} \begin{bmatrix} V^T & U^T \\ V^T & -U^T \end{bmatrix} \right)
$$

## Prob. 14
### (1)
If $A$ is a Hermitian matrix, it satisfies the property $A = A^*$, where $A^*$ denotes the conjugate transpose.
According to the spectral theorem, there exists a unitary matrix $U$ and a diagonal matrix of eigenvalues $\Lambda$ such that $A = U \Lambda U^*$.
The induced $2$-norm of a matrix is defined by the following expression.

$$
\|A\|_2 = \sqrt{\lambda_{max}(A^* A)}
$$

Substituting the identity for Hermitian matrices, we have $A^* A = A^2$.
Using the diagonalization of $A$, we can express $A^2$ in terms of the eigenvalues.

$$
A^2 = (U \Lambda U^*)(U \Lambda U^*) = U \Lambda^2 U^*
$$

The eigenvalues of $A^2$ are exactly the squares of the eigenvalues of $A$, denoted by $\lambda_i^2$.
Consequently, the maximum eigenvalue of $A^* A$ is given by the following.

$$
\lambda_{max}(A^* A) = \max_{i} |\lambda_i|^2 = \left( \max_{i} |\lambda_i| \right)^2 = (\rho(A))^2
$$

Taking the square root of both sides, we reach the desired equality.

$$
\|A\|_2 = \rho(A)
$$

### (2)
Let $\lambda$ be an eigenvalue of $A$ such that its magnitude equals the spectral radius, $|\lambda| = \rho(A)$.
There exists a corresponding non-zero eigenvector $x$ that satisfies the eigenvalue equation.

$$
Ax = \lambda x
$$

By the definition of an induced matrix norm, the following inequality must hold for any vector $x$.

$$
\|Ax\| \leq \|A\| \|x\|
$$

Substituting the eigenvalue equation into the norm on the left-hand side, we get the following.

$$
\|\lambda x\| \leq \|A\| \|x\|
$$

Using the absolute homogeneity property of norms, we pull the scalar $\lambda$ outside.

$$
|\lambda| \|x\| \leq \|A\| \|x\|
$$

Since $x$ is a non-zero vector, we have $\|x\| > 0$, allowing us to divide both sides by $\|x\|$.
This directly yields the inequality for the spectral radius.

$$
\rho(A) \leq \|A\|
$$

### (3)
To prove $\lim_{k \to \infty} \|A^k\|^{1/k} = \rho(A)$, we establish both a lower and an upper bound.
First, we apply the result from the previous section to the matrix $A^k$.

$$
\rho(A^k) \leq \|A^k\|
$$

Since the eigenvalues of $A^k$ are $\lambda_i^k$, it follows that $\rho(A^k) = (\rho(A))^k$.

$$
\rho(A)^k \leq \|A^k\| \implies \rho(A) \leq \|A^k\|^{1/k}
$$

For the upper bound, let $\epsilon > 0$ be an arbitrary constant.
According to the first provided fact, there exists a specific norm $\|\cdot\|_a$ such that $\|A\|_a \leq \rho(A) + \epsilon$.
By the submultiplicative property of induced norms, we have the following bound for $A^k$.

$$
\|A^k\|_a \leq (\|A\|_a)^k \leq (\rho(A) + \epsilon)^k
$$

Using the second fact regarding norm equivalence, there exists a constant $c > 0$ such that our target norm $\|\cdot\|$ is bounded.

$$
\|A^k\| \leq c \|A^k\|_a \leq c (\rho(A) + \epsilon)^k
$$

Taking the $k$-th root of the entire expression, we obtain the following.

$$
\|A^k\|^{1/k} \leq c^{1/k} (\rho(A) + \epsilon)
$$

As $k$ approaches infinity, the term $c^{1/k}$ converges to $1$, leading to the limit bound.

$$
\limsup_{k \to \infty} \|A^k\|^{1/k} \leq \rho(A) + \epsilon
$$

Since this holds for any $\epsilon > 0$, we conclude that the limit equals the spectral radius.

$$
\lim_{k \to \infty} \|A^k\|^{1/k} = \rho(A)
$$

### (4)
Suppose the spectral radius of $A$ is strictly less than $1$.
We can choose a sufficiently small $\epsilon > 0$ such that $\rho(A) + \epsilon = q < 1$.
Based on the facts used in Gelfand's formula, there exists a norm $\|\cdot\|_a$ such that the following is true.

$$
\|A\|_a \leq \rho(A) + \epsilon = q
$$

Considering the norm of the $k$-th power of $A$ in this specific norm space, we have the following inequality.

$$
\|A^k\|_a \leq (\|A\|_a)^k \leq q^k
$$

Because $0 \leq q < 1$, the sequence $q^k$ vanishes as $k$ goes to infinity.

$$
\lim_{k \to \infty} \|A^k\|_a = 0
$$

By the principle of norm equivalence, all norms on a finite-dimensional space are equivalent, including the Frobenius norm.

$$
\|A^k\|_F \leq c \|A^k\|_a
$$

As the right-hand side goes to zero, the Frobenius norm of $A^k$ must also converge to zero.

$$
\lim_{k \to \infty} \|A^k\|_F = 0
$$

Since the convergence of the Frobenius norm to zero implies that every individual entry of the matrix converges to zero, we conclude the final result.

$$
\lim_{k \to \infty} A^k = 0
$$
